---
title: "Differences in Travel Costs in Talbot County, Md between  the White and African American Populations c. 1880"
author: Joe Prego
 
---

# Introduction
This project will examine the differences in spatial distance between the African American and White populations of Talbot County, Maryland during the second half of the 19th century based upon the population listed in the Lake, Griffing and Stevenson 1877 map, “An illustrated atlas of Talbot and Dorchester Counties, Maryland” and the census data from Talbot County in 1880. The Easton Market, located in Easton within greater Talbot County, was the main place to get economic resources during this period. This project will compare the travel distances between the African American and White populations to this market to test for differences in their costs of travel to obtain economic resources. These distances will be In doing so, it will as the questions, did the African American and White populations of Talbot County, Maryland have similar average costs of travel from their dwellings to the Easton town market to obtain economic resources during the later half of the 19th century?

# Materials and methods

The code below will create the raster object that will be used to do the travel cost calculations. It will be a representation of Talbot county with different values assigned to roads, land, and water. The point shapefile will be necessary to get to “to” and “from” points when running the calculations. The gDistance package will be used to generate the least cost paths to the Easton town market for the African American and White population and compare the two. The comparison will be between the average travel costs value generated for each of the populations. In gDistance, this will be done by using the transition object that was created using the raster package which importantly contains the values for moving from cell to cell. The values will essentially be three different categories, with a small value representing roads, a high value representing non-roads, and a higher value representing water. This means, when the test is run, each time it will travel on the least cost path, therefore along the roads, to get from individual houses to the town market.</p>
<p>For analysis, the final package that I will use will be ggplot. The results from the gDistance travel cost tests will need to be analyzed to compare the two populations. A histogram would be the easiest way to represent the results for easy analysis, and it should be accompanied by a table with the numeric results as well.

These packages need to be installed:
```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
library(maptools)
library(sf)
library(raster)
library(rgdal)
library(methods)
library(ggplot2)
library(rasterVis)
library(tidyverse)
```


Then the shapefiles and elevation raster must be read in.
```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}
library(maptools)
library(sf)
library(raster)
library(rgdal)
library(methods)
library(ggplot2)
library(rasterVis)
library(tidyverse)

#This is the road shapefile from the historic map
roads<-read_sf("https://github.com/geo511-2019/2019-geo511-project-jpreg/blob/master/_data/1877_roads_talbot.shp")%>%
  st_transform(4326)

#This is the houses shapefile from the historic map and Census Data
houses<-read_sf("https://github.com/geo511-2019/2019-geo511-project-jpreg/blob/master/_data/1877_houses_talbot.shp")%>%
  st_transform(4326)

#This is the elevation raster of the area, cropped to the extent of the roads shapefile
elev_crop<-raster("https://github.com/geo511-2019/2019-geo511-project-jpreg/blob/master/_data/cropped_elevation_raster_talbot.tif")




```

#Creation of the Project Raster
The code below will create the project raster. This raster combines the elevation tif and road shapefile into a single raster, and assigns values for the land type.  Road is given a value of 1, land is given a value of 4, and water is given a value of 10.  With this information, the gdistance package will eventually be able to calculate travel costs.

```{r, echo=TRUE, warning=FALSE, message=FALSE, error=FALSE}

#This assigns the proper value to water: water
water=(elev_crop<0.001)*6

#Turning the Roads Shapefile into Raster Format
roads_raster <- rasterize(roads, water, field=1, background=4)

#function for values of roads over water: bridges
bridges <- function(x) ifelse (x==13, 1, x)

#Assigning different values to the land cover types
Land_Cover_Guide = c(
  water=10,
  roads=1,
  land=4)

lcd=data.frame(
  ID=Land_Cover_Guide,
  landcover=names(Land_Cover_Guide),
  col=c("blue", "black", "red"),
  stringsAsFactors = F)

talbot_r <- water+roads_raster %>% calc(bridges)

talbot_rf=as.factor(talbot_r)
levels(talbot_rf)=left_join(levels(talbot_rf)[[1]], lcd)

#The code creates a plot of the project raster
roads_water_plot<-gplot(talbot_rf) +
  geom_raster(mapping=aes(fill=value)) +
  coord_cartesian() +
  xlab("longitude") +
  ylab("latitude")
roads_water_plot
  
```


# Results

This is where the results of the gdistance tests will be displayed in table format

# Conclusions


# References
The first data source for this project is the Lake, Griffing and Stevenson 1877 map, "An illustrated atlas of Talbot and Dorchester Counties, Maryland".  This map is available on [Johns Hopkins Library.](https://jscholarship.library.jhu.edu/handle/1774.2/34140) It includes the names of the property owners throughout Talbot County in 1877, along with the locations of their dwelling places.  In addition, the roads are displayed on the map, which will be used to create the travel cost raster necessary for analysis.  The location of the market is determined from another page in this atlas displaying the town of Easton in Talbot County.</p>

<p>The second data source is the 1880 Talbot County Census data, available from [FamilySearch.](https://www.familysearch.org/search/catalog/1417529?availability=Family%20History%20Library) The names from the historic map were paired against the census record to create the necessary dataset that distinguishes between the African American and White population represented on the map.


